---
title: "State of Oregon and Dual Language Immersion"
author: "Manuel Vazquez, Errol Kaylor, Tingyu Zou"
date: "2/25/2022"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme:
      version: 4
      bootswatch: minty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE
                      )
#library(tmap)
library(edld652)
library(tidyverse)
library(skimr)
library(gghighlight)
library(arrow)
library(here)
library(rio)
library(ggrepel)
library(tigris)
library(leaflet)
library(flexdashboard)
library(rgdal)
```

```{r data setup, include=FALSE}

or_district_math_participation <- as_tibble(get_data("EDFacts_math_achievement_lea_2010_2019")) %>% 
  filter(STNAM == "OREGON")

or_district_math_achievement <- as_tibble(get_data("EDFacts_math_participation_lea_2013_2019")) %>% 
  filter(STNAM == "OREGON")

or_district_reading_participation <- as_tibble(get_data("EDFacts_rla_participation_lea_2013_2019")) %>% 
  filter(STNAM == "OREGON")

or_district_reading_achievement <- as_tibble(get_data("EDFacts_rla_achievement_lea_2010_2019")) %>% 
  filter(STNAM == "OREGON")

# or_district_ccd_fiscal_2012 <- as_tibble(get_data("NCES_CCD_fiscal_district_2012")) %>% 
#   filter(STNAM == "OREGON")

or_district_directory <- as_tibble(get_data("NCES_CCD_nonfiscal_district_2017_2021_directory")) %>%
  filter(STATENAME == "OREGON")

or_district_directory <- as_tibble(get_data("NCES_CCD_nonfiscal_district_2017_2021_directory")) %>% 
  filter(STATENAME == "OREGON")

or_district_el <- as_tibble(get_data("NCES_CCD_nonfiscal_district_2017_2021_english_learners")) %>% 
  filter(STATENAME == "OREGON") %>% 
  select(LEAID,
         SCHOOL_YEAR,
         LEP_COUNT)

or_district_membership <- as_tibble(read_parquet(here( "data", "district-membership-17-21.parquet"))) %>% 
  filter(STATENAME == "OREGON")
```


# Oregon District Data

Background {.sidebar data-width=350}
-----------------------------------------------------------------------

The state of Oregon is home to 50,500 students characterized as English Learners across its ~200 districts. These students represent **close to 9%** of reported students in Oregon schools, across all grades. 

Initial Versions of the graph chart EL populations compared to total student population of each district.

Our second visualization plots raw counts of students across Oregon, showing areas of greater density. These counts are not intuitive, and make more sense when displayed relative to the overall number of students. 

Understanding that the relative need for EL student support varies dramatically across Oregon allows for allocating resources to better serve as many students as possible, while acknowledging what more isolated EL students may face.



Column {.tabset}
-----------------------------------------------------------------------

### Chart B

```{r}
or_district_el_directory <- 
 left_join(or_district_directory,
           or_district_el,
           by="LEAID")

or_enroll_total_edit <- or_enroll_total %>% 
  mutate(LEAID = leaid)

full_dataset <- left_join(or_district_el_directory,
                          or_district_reading_achievement,
                          by="LEAID") %>% 
                left_join(or_district_reading_participation,
                          by="LEAID") %>%
                left_join(or_enroll_total_edit,
                          by="LEAID")
or_district_data <- school_districts(state=41)

full_dataset %>% 
  ggplot(aes(student_count,fct_reorder(lea_name,student_count)))+
  geom_point(aes(size=lep_count))
```

### EL Distribution V1

```{r}
pal <- colorNumeric(
 palette = "Blues",
 domain = full_dataset$LEP_COUNT)

first_map <- leaflet(or_district_data) %>%
 addProviderTiles("CartoDB.Positron") %>% 
 addPolygons(color = ~pal(full_dataset$LEP_COUNT),
             weight = 0.5,
             fillOpacity = 1) 

first_map
```



### Chart C

```{r}

```

# Reading + Language Arts Achievement


Column {.tabset}
-----------------------------------------------------------------------
### First Iteration of the graph

```{r }
#Importing the file that has list of districts with dual language immersion schools in Oregon
dli <- import(here("data","districts_dli.xlsx")) %>%
  janitor::clean_names() %>%
  mutate(leaid = as.character(leaid))

# Filtering the EL file to only include years 2018-19
or_el <- or_district_el %>%
  filter(SCHOOL_YEAR == "2018-2019") %>%
  select(LEAID, LEP_COUNT) %>%
  janitor::clean_names()

# From the membership file, taking the number of White Students
or_enroll_white <- or_district_membership %>%
  filter(SCHOOL_YEAR == "2018-2019" & 
           GRADE == "No Category Codes" &
           RACE_ETHNICITY == "White") %>%
  group_by(LEAID) %>%
  summarize(student_white_count = sum(STUDENT_COUNT))

# creating the analytic file with necesarry variables, taking the overall enrollment
# which is represented by the "No category codes" value
or_enroll_total <- or_district_membership %>%
  filter(SCHOOL_YEAR == "2018-2019" & 
           GRADE == "No Category Codes" &
           RACE_ETHNICITY == "No Category Codes" & 
           SEX == "No Category Codes" &
         TOTAL_INDICATOR == "Education Unit Total" &
           STUDENT_COUNT > 0 &
           !is.na(STUDENT_COUNT) ) %>%
  left_join(or_enroll_white, by = "LEAID") %>%
  mutate(white_pct = (student_white_count/STUDENT_COUNT)*100,
         students_color_pct = 100 - white_pct) %>%
  janitor::clean_names() %>%
  left_join(dli,by="leaid") %>%
  left_join(or_el,by = "leaid") %>%
  select(leaid,lea_name,student_count,student_white_count,white_pct,schools_dli, lep_count, students_color_pct) %>%
  mutate(schools_dli = ifelse(is.na(schools_dli),0,schools_dli),
         el_pct = (lep_count/student_count)*100,
         dli_cat = case_when(schools_dli == 0 ~ "0 schools",
                             schools_dli > 0 & schools_dli < 3 ~"1-2 schools",
                             schools_dli > 2 & schools_dli < 6 ~ "3-5 schools",
                             schools_dli > 3 ~"More than 5 schools"
                             ))
  


s <- get_data("EDFacts_rla_achievement_lea_2010_2019")
ach <- filter(s,STNAM=="OREGON") %>% 
  drop_na(LEP_RLA00PCTPROF)

ach2 <- ach %>% 
  rename(leaid=LEAID)

FinalDF <- left_join(or_enroll_total,ach2,by="leaid") %>% 
  drop_na(LEP_RLA00PCTPROF)

#remove rows that are privacy protected
Final <- filter(FinalDF,LEP_RLA00PCTPROF!="PS",
                LEP_RLA00PCTPROF!="GE50",
                LEP_RLA00PCTPROF!="LT50",
                LEP_RLA00PCTPROF!="LE20") %>%
  mutate(value1 = as.numeric(str_extract(LEP_RLA00PCTPROF, "\\d+")),
      value2 = abs(as.numeric(str_extract(LEP_RLA00PCTPROF, "\\-\\d+"))),
      el_prof_read = ifelse(is.na(value2),value1,((value1+value2)/2))) %>%
  select(leaid,lea_name, white_pct,schools_dli,lep_count,students_color_pct,el_pct,dli_cat,LEP_RLA00PCTPROF,value1,value2,el_prof_read) %>%
  arrange(el_prof_read)

dli_read <- Final %>%
  group_by(dli_cat) %>%
  summarize(mean = mean(el_prof_read))

Final %>% 
ggplot(aes(y=forcats::fct_reorder(lea_name,el_prof_read),x=el_prof_read, color = dli_cat)) +
  geom_point(aes(size = el_pct), color = "grey80") +
  geom_point(data = filter(Final,schools_dli > 0 & !is.na(dli_cat)),
             aes(size = el_pct, 
                 color = as.factor(dli_cat)),
                alpha = .7) + 
  geom_vline(xintercept =42, alpha = .6) +
  geom_vline(xintercept =41, linetype = "dashed", color = "red", alpha = .6, size = 1.25) +
  geom_vline(xintercept =44, linetype = "dashed",  color = "green", alpha = .6, size = 1.25) +
  xlim(25,75) +
  theme_minimal() +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 10)) +
  labs(y = " ", x = "Percent proficient in reading")
```

### Graph 2 

```{r}
Final %>% 
  ggplot(aes(el_pct, white_pct)) +
        geom_point(color = "gray70")+
        geom_text_repel(aes(label = LEP_RLA00PCTPROF))+
        theme_minimal() +
        labs(title = "Destrict and Achievement", 
             y = "Percent white", 
             x = "Percent EL of total enrollment")
```


# DLI

Background {.sidebar data-width=350}
-----------------------------------------------------------------------
**Dual language immersion programs** aim to promote high academic achievement, full biliteracy, and develop social consciousness by grouping EL and monolingual students into one classroom while splitting instruction in English and a partner language. Evidence from rigorous studies finds that TWDLI programs has a positive impact on the academic achievement of EL classified students. Leveraging a lottery system to select participants into TWDLI programs at Portland Public Schools ¬– an ideal for an experimental design– Steele et. al., (2017) found that both EL and non-EL students who participated in DLI programs had higher fourth grade English language arts assessments compared to students who were not selected to participate in the program. These results are echoed by quasi-experimental studies in other settings (Watzinger-Tharp et al., 2018) and by meta-analysis that reviewed quantitative studies that interrogated the effects of bilingual education (Rolstad et. al., 2018).

**DLI programs in Oregon are present in at least 24 districts**. Districts with the largest programs tend to be more racially & linguistically diverse. For example, Portland, Beaverton, Hillsboro, Salem-Keizer, and Woodburn all have large DLI programs and serve a high proportion of EL students. However, many of the district programs are in districts that are less diverse.

**There are at least 9 districts in Oregon that serve a high number of EL students but don't offer DLI programs.** These districts serve more than 500 current EL students and have high linguistic and racial diversity. Research indicates that many of these students could benefit from programs such as DLI. 



Column {.tabset}
-----------------------------------------------------------------------

### First Iteration of the graph

```{r }
## Creating a plot that visualizes DLI concentration by white and EL concentration in Oregon
or_enroll_total%>%
  ggplot(aes(x = el_pct, y=white_pct)) +
  geom_point(aes(size = lep_count), color = "grey80") +
  geom_point(data = filter(or_enroll_total,schools_dli > 0 & !is.na(dli_cat)),
             aes(el_pct,y = white_pct, size = lep_count, color = as.factor(dli_cat))) +
  scale_size_continuous(
    name = "Number of EL students",
    breaks = c(50,100,500,750, 1000)) +
  xlim(0,35) +
# create the quadrants
  geom_segment(aes(x = 35, y = 0, xend = 35, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 35, yend = 0)) +
  geom_segment(aes(x = 0, y = 60, xend = 35, yend = 60)) +
  geom_segment(aes(x = 10, y = 0, xend = 10, yend = 100)) +
  geom_segment(aes(x = 0, y = 100, xend = 35, yend = 100)) +
  # quadrant labels
  annotate("text", x = 5, y = 95, alpha = 0.35, label = "High White, Low EL Districts") +
  annotate("text", x = 20, y = 95, alpha = 0.35, label = " ") +
  annotate("text", x = 20, y = 15, alpha = 0.35, label = "Low White, High EL Districts\n\ (Candiates for DLI programs)") +
  annotate("text", x = 5, y = 15, alpha = 0.35, label = " ") +
  theme_minimal() +
  labs(
       y = "Percent white", 
       x = "Percent EL of total enrollment")
  


```

### Final Graph 1. 

```{r }
or_enroll_total%>%
  ggplot(aes(x = el_pct, y=students_color_pct)) +
  geom_point(aes(size = lep_count), color = "grey80") +
  geom_point(data = filter(or_enroll_total,schools_dli > 0 & !is.na(dli_cat)),
             aes(el_pct,y = students_color_pct, 
                 size = lep_count, 
                 color = as.factor(dli_cat)),
             alpha = .7)+
  geom_text_repel(data = filter(or_enroll_total, lep_count > 750 & schools_dli > 5),
                  aes(label = lea_name), min.segment.length = 0, box.padding = 1.05) +
  scale_size_continuous(
    name = "Number of EL students",
    breaks = c(100,500, 1000, 2000, 4000, 6000),
    range = c(1,10)) +
  colorblindr::scale_color_OkabeIto(
    name = "Number of DLI schools"
  ) +
  xlim(0,35) +
# create the quadrants
  geom_segment(aes(x = 35, y = 0, xend = 35, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 35, yend = 0)) +
  geom_segment(aes(x = 0, y = 40, xend = 35, yend = 40)) +
  geom_segment(aes(x = 10, y = 0, xend = 10, yend = 100)) +
  geom_segment(aes(x = 0, y = 100, xend = 35, yend = 100)) +
  # quadrant labels
  annotate("text", x = 5, y = 5, alpha = 0.35, label = " ") +
  annotate("text", x = 20, y = 5, alpha = 0.35, label = " ") +
  annotate("text", x = 20, y = 85, alpha = 0.35, label = "Racially & Linguistically \n\ diverse districts") +
  annotate("text", x = 5, y = 95, alpha = 0.35, label = " ") +
  theme_minimal() +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "Districts with High concentration of EL students have larger DLI programs",
       subtitle = "Yet, there is a high concentration of DLI programs in low diversity districts",
       y = "Percent students of color", 
       x = "Percent EL of total enrollment",
       caption = " \n\ Note: DLI = Dual language immersion") 
```

### Graph 2. Candidates for DLI

```{r }
or_enroll_total%>%
  ggplot(aes(x = el_pct, y=students_color_pct)) +
  geom_point(aes(size = lep_count), color = "grey80") +
  geom_point(data = filter(or_enroll_total,schools_dli > 0 & !is.na(dli_cat)),
             aes(el_pct,y = students_color_pct, 
                 size = lep_count, 
                 color = as.factor(dli_cat)),
             alpha = .7)+
  geom_text_repel(data = filter(or_enroll_total, lep_count > 500 & schools_dli < 1),
                  aes(label = lea_name), min.segment.length = 0, box.padding = 1.05) +
  scale_size_continuous(
    name = "Number of EL students",
    breaks = c(100,500, 1000, 2000, 4000, 6000),
    range = c(1,10)) +
  colorblindr::scale_color_OkabeIto(
    name = "Number of DLI schools"
  ) +
  xlim(0,35) +
# create the quadrants
  geom_segment(aes(x = 35, y = 0, xend = 35, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 100)) +
  geom_segment(aes(x = 0, y = 0, xend = 35, yend = 0)) +
  geom_segment(aes(x = 0, y = 40, xend = 35, yend = 40)) +
  geom_segment(aes(x = 10, y = 0, xend = 10, yend = 100)) +
  geom_segment(aes(x = 0, y = 100, xend = 35, yend = 100)) +
  # quadrant labels
  annotate("text", x = 5, y = 5, alpha = 0.35, label = " ") +
  annotate("text", x = 20, y = 5, alpha = 0.35, label = " ") +
  annotate("text", x = 20, y = 85, alpha = 0.35, label = "Linguistically diverse districts \n\ (Candidates for DLI programs)") +
  annotate("text", x = 5, y = 95, alpha = 0.35, label = " ") +
  theme_minimal() +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "In 2018-19, at least nine Oregon Districts served more than 500 EL students and did not offer DLI",
       subtitle = "",
       y = "Percent students of color", 
       x = "Percent EL of total enrollment",
       caption = " \n\ Note: DLI = Dual language immersion") 
```

